A Graph Theoretic Approach to Problem Formulation for Multidisciplinary Design Analysis and Optimization

### 2013-05-09 intro.tex
removing old Fig 1.
update

    Our goal is to develop formalism for expressing analysis interconnectivity and for determining feasible
    data flows to assist an engineering team conducting this task. Because the problem deals with
    interconnectivity, we base our approach on the representations and techniques of graph theory. The 
    proposed graph syntax is closely related to the REMS system proposed by 
    Alexandrov and Lewis\cite{alexandrov2004}. The approach taken here should be viewed as an extention 
    of the concepts proposed in REMS with the goal of allowing the graphs to more closely interact with 
    existing design processes and MDAO frameworks.

The approach begins by 
    constructing the \emph{maximal connectivity graph (MCG)} describing all possible
    interconnections between the analysis tools proposed by the engineers. Graph operations are then
    conducted to reduce the MCG down to a \emph{fundamental problem graph (FPG)} that describes the set of analysis
    tools needed to solve the specified system-level design problem.
    Any particular solution procedure for the problem; any relevant MDO solution architecture, e.g. MDF,
    IDF, CO, BLISS, could be selected \emph{post facto} to implement the system-level optimization by
    constructing a \emph{problem solution graph (PSG)} and the corresponding problem solution approach.
    The concept of the FPG and the identification of feasible FPGs are the main contributions of the paper.

@@

\section{Motivation}
\subsection{Specific vs Fundamental Problem Formulation }
	\label{s:specific vs fundamental}
    The Fundamental Problem Formulation (FPF) is our terminology for a 
    statement of the overall system-level optimization problem that does not 
    depend on the choice of solution-specifc elements such as the MDAO 
    framework, optimization algorithm, iterative solver, or execution sequence
    used to solve the problem. The FPF contrasts with problem statements that are written without 
    reference to a specific solution strategy, but that still imply a specific execution sequence.
    For example, consider the Sellar test problem \cite{AIAA:sellar}, with an FPF given as follows:

@@

    In Eq. \ref{eqn:simple_fpf}, $y_1$ is an output of $D_1$ as well as an input 
    to $D_2$. Similarly, $y_2$ is an output of $D_2$ as well as an input 
    to $D_1$. This implies that $D_1$ and $D_2$ are coupled together by means 
    of their reciprical dependence. However no information is given as to weather
    $D_1$ or $D_2$ should be run before the other. Given Eq \ref{eqn:simple_fpf}, 
    it would be equally valid to run $D_1$ first, $D_2$ first, or both simultaneously.

    Eq. \ref{eqn:simple} gives a modified problem formulation where $D_1$ is run 
    first. A new variable is introduced, $\hat{y_2}$, to break the dependence of
    $D_1$ on $D_2$. Also a new coupling constraint, $H$, to enforce the consistency
    that would otherwise be lost due to $\hat{y_2}$. Equation \ref{eqn:simple} is 
    a specific alternative expression of the more fundamental problem description 
    given in Eq. \ref{eqn:simple_fpf}.

@@

    Figure \ref{f:sellar graph} depicts the Sellar problem given by 
    Eq.~\ref{eqn:simple_fpf} in the graph-based syntax which will be presented 
    in this paper. This example will serve to highlight the features of the 
    syntax as they are introduced.
    \begin{figure}[htb!]
      \begin{center}
        \includegraphics[width=4.0in]{images/sellar}
      \end{center}
      \caption{Sellar problem represented as a graph.}
    \label{f:sellar graph}
    \end{figure}

@@

    A DSM by itself is insufficient to describe complete problem formulations because it 
    captures only information about data dependency between analyses; objective and 
    constraint information is missing from the DSM description of the problem. 
    An alternative matrix-based syntax, called a Functional Dependency Table (FDT), was proposed by Wagner and Papalambros. 
    FDT represents the relationship between functions, including objectives and constraints, and specific variables that affect 
    them\cite{Wagner1993}. Similar to DSM, FDT also describes an adjacency matrix of a graph. Unlike the DSM graph, 
    however, the FDT graph is undirected and nodes can represent analysis tools, objectives, 
    or constraints. Edges between nodes represent a dependence on the same 
    variable. Michelena and Papalambros made use of the FDT to solve a graph partitioning problem that yielded 
    more efficient optimization problem decompositions\cite{Michelena1997}. While FDT succeeds at capturing the 
    information about objectives and constraints, its lack of directed edges 
    implies that it cannot capture the coupled data dependency that a DSM captures. 
    The FDT in Fig. \ref{fig:FDT_simple} shows that $D_1$ is dependent on $y_2$ and 
    that $D_2$ is dependent on $y_1$, but you can not infer their coupled dependence from 
    that information alone. This missing information implies 
    that, while FDT is very useful for partitioning problems, it is also not 
    sufficient to completely describe a problem formulation. 
    \begin{figure}[htb!]
        \begin{center}
        \begin{tabular}{|c|c|c|c|c|c|c|}
            \hline
                   & $x_1$ & $y_1$ & $y_2$ & $z_1$ & $z_2$ \\ \hline
            $D_1$  & 1     &       & 1     & 1     & 1     \\ \hline
            $D_2$  &       & 1     &       & 1     & 1     \\ \hline
            $F$    & 1     & 1     & 1     &       & 1     \\ \hline
            $G_1$  &       & 1     &       &       &       \\ \hline
            $G_2$  &       &       & 1     &       &       \\
            \hline
        \end{tabular}
        \caption{Functional Dependency Table (FDT) for Eq. \ref{eqn:simple_fpf} \label{fig:FDT_simple}}
        \end{center}
    \end{figure}

    Alexandrov and Lewis introduced a graph based syntax called REMS which 
    incorporates objectives and constraints into a graph, effectively combining 
    FDT and DSM\cite{alexandrov2004}. REMS retains the square adjacency 
    matrix form from DSM, but by adding the objectives and constraints, it partially 
    combines a traditional DSM with an FDT. This formulation allows REMS to represent data 
    dependency between multiple analysis tools as well as between analysis tools and
    objective/constraint functions. Additionaly, REMS address the need to be able to 
    transition between multiple solution strategies while keeping a constant 
    graph representation of the fundamental problem formulation. REMS provides
    syntax for variables and functions but does not facilitate inclusion 
    of solvers or optimizers in the graph.

@@

\subsection{Requirements for a New Graph Syntax}
  \label{s:requirements}
  The goal of the new graph-based syntax presented here is to enable the general structure of an MDAO problem to be described independently of any solution information, 
  while still being able to accomodate the more specific case when a solution 
  strategy is applied. In order to achieve this goal, 
  the graph syntax needs to accomodate a number of MDAO problem constructs: 
  \begin{itemize}
    \item Analysis tools and their interconnections
    \item Design variables, objectives, and constraints
    \item Local and global information
    \item Coupling between analyses
    \item Multi-fidelity analyses
  \end{itemize}

>>>>

    Our goal is to develop formalism for expressing analysis interconnectivity and for determining feasible
    analysis tool sets to assist an engineering team conducting this task. Because the problem deals with
    interconnectivity, we base our approach on the representations and techniques of graph theory.
    The approach begins by constructing the \emph{maximal connectivity graph (MCG)} describing all possible
    interconnections between the analysis tools proposed by the engineers. Graph operations are then
    conducted to reduce the MCG down to a \emph{fundamental problem graph (FPG)} that describes the set of analysis
    tools needed to solve the specified system-level design problem.
    Any particular solution procedure for the problem; any relevant MDO solution architecture, e.g. MDF,
    IDF, CO, BLISS, could be selected \emph{post facto} to implement the system-level optimization by
    constructing a \emph{problem solution graph (PSG)} and the corresponding problem solution approach.
    The concept of the FPG and the identification of feasible FPGs are the main contributions of the paper.

@@

\section{Motivation}
\subsection{Specific vs Fundamental Problem Formulation }
	\label{s:specific vs fundamental}
    The Fundamental Problem Formulation (FPF) is our terminology for a 
    statement of the overall system-level optimization problem that contains only 
    information about analysis tools, design variables, objectives, and constraints without any 
    information about how to solve the problem. It does not depend on the choice 
    of solution-specific elements such as the MDAO framework, optimization 
    algorithm, iterative solver, or execution sequence. This description represents 
    the problem from the point of view of a engineer specifying the details of a design problem 
    without constrainig how it should be solved. The lack of solution information 
    in the FPF contrasts with problem statements that are written with reference 
    to a specific solution strategy or with implication of a specific execution sequence.
    For example, consider the Sellar test problem \cite{AIAA:sellar}, with an FPF given as follows:

@@

    In Eq.~\ref{eqn:simple_fpf}, $y_1$ is an output of $D_1$ as well as an input 
    to $D_2$. Similarly, $y_2$ is an output of $D_2$ as well as an input 
    to $D_1$. This implies that $D_1$ and $D_2$ are coupled together by means 
    of their reciprical dependence. However no information is given as to weather
    $D_1$ or $D_2$ should be run before the other. Given Eq \ref{eqn:simple_fpf}, 
    it would be equally valid to run $D_1$ first, $D_2$ first, or both simultaneously.

    The Sellar problem is very simple. There is only a single coupling interaction 
    between the two disciplines and there are a very limited set of variables. 
    That makes it relatively easy to describe the problem given in 
    Eq.~\ref{eqn:simple_fpf}. However, for larger more complex problems, 
    it is much harder to identify the fundamental problem formulation. Interdisciplinary 
    couplings are not always obvious and the set of analsyis tools and variables are 
    much larger. To further compount the problem engineers may already have some 
    solution strategy or execution order they are used to applying to problems. 
    Hence, it is common to see problem formulations given in a less general form 
    as in Eq. \ref{eqn:simple}. Here $D_1$ must be run 
    first. A new variable is introduced, $\hat{y_2}$, to break the dependence of
    $D_1$ on $D_2$. Also a new coupling constraint, $H$, is added to enforce the 
    consistency that would otherwise be lost due to $\hat{y_2}$. Equation \ref{eqn:simple} is 
    a more specific, but equally valid, expression of the Sellar problem that could easily be 
    generated through an incremental approach to constructing the problem.

@@

@@

    A DSM by itself is insufficient to describe complete problem formulations because it 
    captures only information about data dependency between analyses; objective and 
    constraint information is missing from the DSM description of the problem. 
    An alternative matrix-based syntax, called a Functional Dependency Table (FDT), was proposed by Wagner and Papalambros. 
    FDT represents the relationship between functions, including objectives and constraints, and specific variables that affect 
    them\cite{Wagner1993}. Similar to DSM, FDT also describes an adjacency matrix of a graph. Unlike the DSM graph, 
    however, the FDT graph is undirected and nodes can represent analysis tools, objectives, 
    or constraints. Edges between nodes represent a dependence on the same 
    variable. Michelena and Papalambros made use of the FDT to solve a graph partitioning problem that yielded 
    more efficient optimization problem decompositions\cite{Michelena1997}. Allison 
    extended this work to couple adjacency matrix information with the FDT to 
    account for system coupling in an automated partitioning scheme\cite{Allison2008} While FDT succeeds at capturing the 
    information about objectives and constraints, its lack of directed edges 
    implies that it cannot capture the coupled data dependency that a DSM captures. 
    The FDT in Fig. \ref{fig:FDT_simple} shows that $D_1$ is dependent on $y_2$ and 
    that $D_2$ is dependent on $y_1$, but you can not infer their coupled dependence from 
    that information alone. This missing information implies 
    that, while FDT is very useful for partitioning problems, it is also not 
    sufficient to completely describe a problem formulation.

    \begin{figure}[htb!]
        \begin{center}
        \begin{tabular}{|c|c|c|c|c|c|c|}
            \hline
                   & $x_1$ & $y_1$ & $y_2$ & $z_1$ & $z_2$ \\ \hline
            $D_1$  & 1     &       & 1     & 1     & 1     \\ \hline
            $D_2$  &       & 1     &       & 1     & 1     \\ \hline
            $F$    & 1     & 1     & 1     &       & 1     \\ \hline
            $G_1$  &       & 1     &       &       &       \\ \hline
            $G_2$  &       &       & 1     &       &       \\
            \hline
        \end{tabular}
        \caption{Functional Dependency Table (FDT) for Eq. \ref{eqn:simple_fpf} \label{fig:FDT_simple}}
        \end{center}
    \end{figure}

    Alexandrov and Lewis introduced a graph based syntax called REMS which 
    incorporates objectives and constraints into a graph, effectively combining 
    FDT and DSM\cite{alexandrov2004}. REMS retains the square adjacency 
    matrix form from DSM, but by adding the objectives and constraints, it partially 
    combines a traditional DSM with an FDT. This formulation allows REMS to represent data 
    dependency between multiple analysis tools as well as between analysis tools and
    objective/constraint functions. Additionaly, REMS address the need to be able to 
    transition between multiple solution strategies while keeping a constant 
    graph representation of the fundamental problem formulation. REMS provides
    syntax for variables and functions but does not facilitate inclusion 
    of solvers or optimizers in the graph.  The 
    proposed graph syntax is closely related to REMS. The syntax proposed here should 
    be viewed as an extention to REMS with the goal of allowing the graphs to more 
    closely interact with existing design processes and MDAO frameworks.

@@

\subsection{Requirements for a New Graph Syntax}
  \label{s:requirements}
  The goal of the new graph-based syntax presented here is to enable the general 
  structure of an MDAO problem to be described independently of any solution information, 
  while still being able to accomodate the more specific case when a solution 
  strategy is applied. In order to achieve this goal, 
  the graph syntax needs to accomodate a number of MDAO problem constructs: 
  \begin{itemize}
    \item Analysis tools and their interconnections
    \item Design variables, objectives, and constraints
    \item Coupling between analyses
    \item Multi-fidelity analyses
  \end{itemize}
